Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase Repair Time?

### 2015-03-12 data.tex
added scatter plots of phase delay and other charts.

We examine the phase delay effect bin
171 software projects from around the world from 
\bill{2005 to 2013} that used the Team Software Process (TSP).

@@

\sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}

Three common features of all those  are (1)~coaches,
(2)~peer review, and  (3)~personnel reviews.

A ``coach'' is a team member  authorized to submit project data
Before submission,
these coaches check the data for obvious errors.

``Personnel review'' is a technique taken  from the Personal Software
ProcessSM (PSPSM). PSPSM encourages developers to continually make and review their personnel estimates
about their day-to-day tasks, then compare those estimates against the actual development effort.
In this way, developers can acquire a more realistic understanding of their work behaviour.

``Peer review'' is a  technique in
traditional software engineering.
 Basili and Boehm write  commented in 2001~\cite{boehm01} 
that peer reviews can catch over half the defects introduced into a system.
Peer review can be conducted on any artifact generated anywhere in the software
lifecycle and can quickly be adapted to new kinds of artifacts.

@@

To aid in monitoring these 171 projects  Nichols  built a
data collection tool that gathers project data a central database at SEI. 
The issue types collected in the the SEI TSP data set divide into:
\be 
\item Environment: design, compile, test, or other support system problems;
\item Interface: procedure calls and reference, I/O, user format;
\item Data: structure, content; 
\item Documentation: comments, messages;
\item Syntax: spelling, punctuation typos, instruction formats
\item Function: logic, pointers, loops, recursion, computation, function defects  
\item Checking: error messages, inadequate checks;
\item Build: change management, library, version control;
\item Assignment: package
declaration, duplicate names, scope, limits;
\item System: configuration, timing, memory.
\ee
The tool also collected process information about those defects.
That data includes  work start time, work end time, delta
work time, and interruption time. Software engineers are often
interrupted by meetings, requests for technical help, reporting, and
so forth. These events are recorded, in minutes, as interruption
time. In this paper, when we report ``time to resolve an
issue'', we show the difference between the start and end times
of a work session, with any interruption time subtracted (the
difference in times, minus the interruptions).

As of January 2014, the SEI TSP database contained data from 109
TSP projects. The projects started between July 2009 and
September 2013; they included 34 teams and 309 people. Among
the database fact tables, the time store contains 103,023 time logs,
18,408 defect logs, and 7,464 size logs\footnote{\bill{we need to 
update this table}}.

@@

\subsection{Project Phases}
This paper studies the impact of phase delay on the time required to resolve issues.
The phased used in this paper are shown in \fig{waterfall}. Note that, in that figure:
\bi 
\item
Several  phases have the same  sub-activities of {\em review} and {\em inspect}\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
\item Testing is divided into several statges \footnote{\bill{need a one
line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
\ei

@@

TSP is not antithetical to agile--  indeed a TSP-waterfall style project can adopt aspects
of agile.  For example the agile loop  could be applied
over one or more of the phases shown in the long waterfall chain of \fig{waterfall}. Some
TSP teams adopt something like test-driven-development\footnote{TDD is an agile-method
where the ``test'' is the primary driver of the design. The tests are written first,
then the code to support those tests. TDD proceeds in three steps: red (where there
are broken tests); green (where tests are passing); and, possibly, refactor (where
the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
reviews are scheduled after testing. Some of the groups in the SEI TSP data used that approach but
they do not effect the main conclusions of this paper:
\bi 
\item They were strongly in the minority\bill{any numbers on this?};
\item Most of our phase delay data comes from much early in the waterfall model
of \fig{waterfall}.
\ei 
That said, there are some very ``un-agile'' aspects
of the processes used by the   projects in the TSP SEI data.
A TSP participant spends much time reflecting on the project,
undisturbed by the behaviour of the  executables.
In fact,
it can be days/weeks
before TSP participants gain  feedback from executing code, for the following reasons:  
\bi 
\item
For TSP projects, the times spent in the of a design activity/phase is   
approximately as much effort as the coding phase.  
\item Our TSP teams did not
 combine development and  testing. 
 \item  The TSP projects studied here offer new releases at least every three months, but
often much longer that that\bill{any numbers on that}. Note that this is a far slower
release schedule than seen in, say, continuous integration projects.  
\item The time devoted to personnel review'' and peer review (defined above)
is  about 50\% as much effort as the previous construction phase activity.

>>>>

We examine the phase delay effect bin
212 software projects from around the world from 
\bill{2005 to 2015} that used the Team Software Process (TSP).

@@

\sei{One paragraph summary on the key points of TSP execution? The following text should also be reviewed and better integrated.}
TSP guiding principles include 1) engineering work does not happen by accident, it must be planned, 2) the plans must include realistic commitments of the resources needed to perform the work, 3) planed and actual results should be frequently examined to assure that plans remain relevant, and 4) people performing the work should be responsible for planning and tracking the work. After clarifying goals, defining work practices, and estimating size and effort, developers make a detailed short term plan. As the developers perform project work, they use a tool such as the Process Dashboard to collect their time, effort, size, and schedule data. Every week, the team review their data to evaluate status, review actual rates, determine if project goals for schedule, cost, and quality are being met and then make necessary plan corrections. At the end of the project the coach and team perform a quantitative post mortem.

Four common features of TSP projects include (1)~coaches,
(2)~peer review, and  (3)~personnel reviews (4)~planning.

A ``coach'' works with the team and is authorized to submit project data.
Before reviewing data with the teams, therefore before submission,
these coaches check the data for obvious errors.

``Personnel review'' is a technique taken  from the Personal Software
ProcessSM (PSPSM) and is more or less unique in its TSP implementation. The technique of a personal review is similar to that of a peer review. Developers systematically examine their own work products using a checklist built from their personal defect profile. This personal review occurs after a portion of a product is done and before peer reviews.

%PSPSM teaches developers howto continually make and review their personnel estimates
%about their day-to-day tasks, then compare those estimates against the actual development effort.
%In this way, developers can acquire a more realistic understanding of their work behaviour.

``Peer review'' is a  technique in
traditional software engineering.
 Basili and Boehm write  commented in 2001~\cite{boehm01} 
that peer reviews can catch over half the defects introduced into a system.
Peer review can be conducted on any artifact generated anywhere in the software
lifecycle and can quickly be adapted to new kinds of artifacts. TSP peer reviews follow the Fagan style in which the reviewer uses a checklist composed of common team defects prior to a review team meeting.

''Planning '' is another technique from the in the PSP(SM). Developers estimate the size of work products, define the process phases, use historic rates and time in phase to estimate the required time, the number of defects injected, and the defects to be removed in each phase. Coaches help the developers to compare estimates against actual results. In this way, developers acquire a more realistic understanding of their work behavior and performance.

@@

\sei{ wrn: please review the following}

Organizations using TSP agree to provide project data to the SEI for use in research with the provision that the data must not be traceable to its source. The data includes project and site  characteristic summaries, surveys of team members, launch presentations, launch outbriefs, and baseline plans, final data from the project, and a post mortem report.  In practice, this data requirement has only been enforced for the purposes of certifying and reauthorizing TSP coaches who are the only individuals authorized to submit data. To this data, only the data recorded using the Process Dashboard tool has been collected and aggregated for research.

To aid in using this data for research and benchmarking  Nichols first collected the Process Dashboard data from the submission repository then used a tool built by the Process Dashboard developer, David Tuma, 
to gather the project data into a database at SEI. Nichols and Shirai then extracted data into views suitable for analysis. . The key views include the a project results summary, effort logs, task logs, and defect logs. At this stage, the data has not yet undergone additional screening.

\sei{
do we need the following information? we do not use it in this analysis
}

The issue types collected in the the SEI TSP data set divide into:
\be 
\item Environment: design, compile, test, or other support system problems;
\item Interface: procedure calls and reference, I/O, user format;
\item Data: structure, content; 
\item Documentation: comments, messages;
\item Syntax: spelling, punctuation typos, instruction formats
\item Function: logic, pointers, loops, recursion, computation, function defects  
\item Checking: error messages, inadequate checks;
\item Build: change management, library, version control;
\item Assignment: package
declaration, duplicate names, scope, limits;
\item System: configuration, timing, memory.
\ee
The tool also collected process information about those defects.
That data includes  work start time, work end time, delta
work time, and interruption time. Software engineers are often
interrupted by meetings, requests for technical help, reporting, and
so forth. These events are recorded, in minutes, as interruption
time. In this paper, when we report ``time to resolve an
issue'', we show the difference between the start and end times
of a work session, with any interruption time subtracted (the
difference in times, minus the interruptions).

%wrn, updated 3/11 these
As of February 2015, the SEI TSP database contained data from 212
TSP projects. The projects completed between July 2008 and
November 2014; they included 47 organizations and 843 people. Among
the database fact tables, 
contains 268,726 time logs, 
154,238 task logs,
 47,376 defect logs, 
and 26,534 size logs
%\footnote{\bill{we need toupdate this table}}.

@@

\subsection{Project Phases}
This paper studies the impact of phase delay on the time required to resolve issues.
The logical phases used in this paper are shown in \fig{waterfall}. Note that, in that figure:
\bi 
\item
Several  phases have the similar  sub-phases of {\em review} and {\em inspect}. In TSP, individuals perform personal reviews on products prior to the peer inspection.\footnote{\bill{plz distinguish review and inspect. does one usually happen first?}}
\item Testing is divided into several statges \footnote{\bill{
(this appears to be OBE), need a one
line description of test vs qualtest vs inttest vs systemTest vs AcceptTest. I tries some words in the Key to \fig{sedata}. dont know if i got it right.}}
\ei

@@

TSP is not antithetical to agile--  indeed a TSP-waterfall style project can adopt aspects
of agile.  For example the agile loop  could be applied
over one or more of the phases shown in the long waterfall chain of \fig{waterfall}. Some
TSP teams adopt something like test-driven-development\footnote{TDD is an agile-method
where the ``test'' is the primary driver of the design. The tests are written first,
then the code to support those tests. TDD proceeds in three steps: red (where there
are broken tests); green (where tests are passing); and, possibly, refactor (where
the code is re-organized based on feedback from those tests~\cite{fraser03}.} where
code inspections are scheduled after unit testing. Some of the groups in the SEI TSP data used that approach but
they do not effect the main conclusions of this paper:
\bi 
\item They were less than 5\% of the total \bill{any numbers on this?};

\item Most of our phase delay data comes from much early in the waterfall model
of \fig{waterfall}.
\ei 
That said, there are some very ``un-agile'' aspects
of the processes used by the   projects in the TSP SEI data.
A TSP participant spends much time reflecting on the project,
undisturbed by the behaviour of the  executables.
In fact,
it can be days/weeks
before TSP participants gain  feedback from executing code, for the following reasons:  
\bi 
\item
For TSP projects, the times spent in the of a design activity/phase is   
approximately as much effort as the coding phase.  
\item Our TSP teams did not
 combine development and  testing. 
 \item  The TSP projects studied here typically offer new releases at least every three months, but
often much longer that that\bill{any numbers on that}. Note that this is a far slower
release schedule than seen in, say, continuous integration projects.  
\item The time devoted to personnel review'' and peer review (defined above)
is  about 50\% as much effort as the previous construction phase activity.

### 2015-03-12 phaseDelay.tex
added scatter plots of phase delay and other charts.

\usepackage{picture}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}} 
\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} \newcommand{\tim}[1]{\textcolor{orange}{TIM: #1}}
\newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
\newcommand{\carter}[1]{\textcolor{cyan}{CARTER: #1}} 
\newcommand{\sei}[1]{\textcolor{RedViolet}{BILL/FORREST: #1}} 
\newcommand{\todo}[1]{\textcolor{Maroon}{TODO: #1}} 
%\newenvironment{changed}{\par}{\par}

@@

\conferenceinfo{FSE}{'15 Bergamo, Italy}
\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increases  Repair Time?}
\numberofauthors{3}
\author{
\alignauthor
Tim Menzies, \\Carter Pape\\
       \affaddr{CS, NcState, USA}\\
       tim.menzies@gmail.com,\\carterpape@gmail.com
\alignauthor
William Nichols,\\ Forrest Schull\\
\affaddr{SEI, CMU, USA}
wrn,fjshull@sei.cmu.edu
\alignauthor
Lucas \\Layman\\
       \affaddr{Fraunhofer Center,  USA}\\ 
       llayman@fc-md.umd.edu
}

@@

This paper shows that this belief is maintained
quite strongly in the industrial software engineering
community (but, in the academic community, somewhat less so).
Yet based on a sample of 
171 software projects from around the world from 
\bill{2005 to 2013}, we argue that this belief is mostly 
incorrect. Specifically: the median fix time for issues
is just a few minutes and this does not tend to increase
the longer the issue remains in the system.

@@

The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}. \todo{Overlay median on Boehm's chart \fig{cost-to-fix}. How does it compare?}

@@

\begin{figure}[!ht]
\begin{center}
\begin{tabular}{r|rrr|l}
  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
size & 25th & 50th & 75th & Phase removed \\
\hline
120 & 3 & 3 & 18 & ReqtsInspect.dat \\
126 & 3 & 3 & 20 & ReqtsReview.dat \\
70 & 2 & 2 & 10 & HLDInspect.dat \\
 55 &    2 &    1&    7&HLDReview.dat \\

278&    7&    7&   33&DesignInspect.dat \\
233&    6&    6&   25&DesignReview.dat \\

315&    8&    7&   34&CodeInspect.dat \\ 
282&    7&    7&   32&CodeReview.dat\\

425&   11&   10&   54&UnitTest.dat \\ 
226&    7&    6&   46&IntTest.dat \\ 
194&    7&    6&   52&SysTest.dat\\ 
 88&    4&    3&   38&AcceptTest.dat\\
\end{tabular}
\end{center}
\caption{Median fix time per phase}
\label{fig:fix-time-per-phase}
\end{figure}

@@

In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected). The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (Table~\ref{tab:phase_corr}).

@@

@@

\begin{table}[ht]
\centering
\begin{tabular}{lll}
  Phase & $tau$ & $p$ \\ 
  \hline
BeforeDevelopment & NA & -- \\ 
  Planning & NA & -- \\ 
  Reqts & NA & -- \\ 
  ReqtsReview & 0.05 & 0.01 \\ 
  ReqtsInspect & 0.17 & $<$0.01 \\ 
  HLD & NA & -- \\ 
  HLDReview & 0.04 & 0.24 \\ 
  HLDInspect & -0.07 & 0.05 \\ 
  Design & 0.19 & 0.08 \\ 
  DesignReview & 0.08 & $<$0.01 \\ 
  DesignInspect & 0.06 & $<$0.01 \\ 
  Code & 0.21 & $<$0.01 \\ 
  CodeReview & 0.15 & $<$0.01 \\ 
  CodeInspect & 0.11 & $<$0.01 \\ 
  Test & 0.16 & $<$0.01 \\ 
  QualTest & NA & -- \\ 
  IntTest & 0.13 & $<$0.01 \\ 
  SysTest & 0.32 & $<$0.01 \\ 
  AcceptTest & 0.07 & 0.29 \\ 
  \end{tabular}
\caption{Correlation of cost-to-fix and phase delay for defects found in each phase (Kendall's $\tau$)} 
\label{tab:phase_corr}
\end{table}

>>>>

\usepackage{subfigure}
\usepackage{picture}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}} 
\newcommand{\lucas}[1]{\textcolor{red}{LUCAS: #1}} \newcommand{\tim}[1]{\textcolor{orange}{TIM: #1}}
\newcommand{\bill}[1]{\textcolor{blue}{BILL: #1}} 
\newcommand{\carter}[1]{\textcolor{cyan}{CARTER: #1}} 
\newcommand{\sei}[1]{\textcolor{RedViolet}{BILL/FORREST: #1}} 
\newcommand{\todo}[1]{\textcolor{Maroon}{TODO: #1}} 
%\newenvironment{changed}{\par}{\par}

@@

\conferenceinfo{FSE}{'15 Bergamo, Italy}
\title{Revisiting the Truisms of Software Engineering:\\ Does Phase Delay Dramatically Increase  Repair Time?}
\numberofauthors{3}
\author{
\alignauthor
Tim Menzies, \\Carter Pape\\
       \affaddr{CS, NcState, USA}\\
       tim.menzies@gmail.com,\\carterpape@gmail.com
\alignauthor
William Nichols,\\ Forrest Schull\\
\affaddr{SEI, CMU, USA}
wrn,fjshull@sei.cmu.edu
\alignauthor
Lucas \\Layman\\
       \affaddr{Fraunhofer Center,  USA}\\ 
       llayman@fc-md.umd.edu
}

@@

This paper shows that this belief is maintained
quite strongly in the industrial software engineering
community (but, in the academic community, somewhat less so).
Yet based on a sample of 
212 software projects from around the world from 
\bill{2005 to 2014}, we argue that this belief is mostly 
incorrect. Specifically: the median fix time for issues
is just a few minutes and this does not tend to increase
the longer the issue remains in the system.

@@

The median time to fix a defect in each phase is shown in \fig{fix-time-per-phase}. The median cost to fix a defect is low in the early phases, and then increases slightly during design and coding before peaking during the unit test phase. The median fix time then drops during the testing phases. Note that while the median time remains relatively consistent, the 75th percentile does increase during the testing phases.

@@

\begin{figure}[!ht]
\begin{center}
\begin{tabular}{r|rrr|l}
  Sample&\multicolumn{3}{c|}{Percentiles}\\ 
size & 25th & 50th & 75th & Phase removed \\
\hline
120 & 3 & 3 & 18 & ReqtsInspect \\
126 & 3 & 3 & 20 & ReqtsReview \\
70 & 2 & 2 & 10 & HLDInspect \\
 55 &    2 &    1&    7&HLDReview \\

278&    7&    7&   33&DesignInspect \\
233&    6&    6&   25&DesignReview \\

315&    8&    7&   34&CodeInspect \\ 
282&    7&    7&   32&CodeReview\\

425&   11&   10&   54&UnitTest \\ 
226&    7&    6&   46&IntTest \\ 
194&    7&    6&   52&SysTest\\ 
 88&    4&    3&   38&AcceptTest\\
\end{tabular}
\end{center}
\caption{Median fix time per phase}
\label{fig:fix-time-per-phase}
\end{figure}

The relative cost to fix errors per lifecycle phase in the TSP project data is overlayed with Boehm's exponential cost curve~\cite{Boehm81} as well as the Royce case study~\cite{Royce98} in \fig{cost-to-fix-tsp}.

\begin{figure}[!ht]
 \includegraphics[width=3.3in]{boehm_overlay.png}
 \caption{Cost-to-fix curve from \fig{cost-to-fix} overlayed with TSP data and case study from~\cite{Royce98}.}
 \label{fig:cost-to-fix-tsp}
 \end{figure}

@@

 \fig{phase_fixtime_plot} shows scatterplots of the cost-to-fix (rank) per phase against the phase delay. In the early phases, i.e., Design \& Coding, most of the defects found have a relatively small phase delay. This is ideal considering that the purpose of the early phase reviews is to remove as many defects as possible.

In the TSP data, there is a weak correlation (Kendall's $\tau = 0.19, p < 0.01, 95\% CI$) between a defect's cost-to-fix (minutes of effort) and the defects phase delay (phase removed - phase injected) - see \fig{correlation}. The correlations were also calculated for each phase, e.g., does the phase delay of defects found in Test correlate with the defect's cost-to-fix (\fig{phase_corr}).

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=3.3in]{corr.png}
\end{center}
\caption{Correlation of phase delay and fix time (plot shows a 1\% sampling of data).}
\label{fig:correlation}
\end{figure}

\begin{figure}[ht]
\centering
\begin{tabular}{lll}
  Phase & $tau$ & $p$ \\ 
  \hline
  Design & 0.14 & 0.08 \\ 
  DesignReview & 0.06 & $<$0.01 \\ 
  DesignInspect & 0.05 & $<$0.01 \\ 
  Code & 0.27 & $<$0.01 \\ 
  CodeReview & 0.12 & $<$0.01 \\ 
  CodeInspect & 0.09 & $<$0.01 \\ 
  Test & 0.13 & $<$0.01 \\ 
  IntTest & 0.1 & $<$0.01 \\ 
  SysTest & 0.25 & $<$0.01 \\ 
  AcceptTest & 0.06 & 0.29 \\ 
  \end{tabular}
\caption{Correlation of cost-to-fix and phase delay for defects found (Kendall's $\tau$)} 
\label{fig:phase_corr}
\end{figure}

\begin{figure*}[!ht]
\centering
\includegraphics[width=1.6in]{corr-Code.png}
\includegraphics[width=1.6in]{corr-CodeReview.png}
\includegraphics[width=1.6in]{corr-CodeInspect.png}
\vspace{0.01cm}

\includegraphics[width=1.6in]{corr-Design.png}
\includegraphics[width=1.6in]{corr-DesignReview.png}
\includegraphics[width=1.6in]{corr-DesignInspect.png}
\vspace{0.01cm}

\includegraphics[width=1.6in]{corr-Test.png}
\includegraphics[width=1.6in]{corr-IntTest.png}
\includegraphics[width=1.6in]{corr-SysTest.png}
\includegraphics[width=1.6in]{corr-AcceptTest.png}

\caption{Correlation of cost-to-fix (rank) and phase delay for defects found by phase for phases where \# defects >= 30 and \# unique phase delay values >= 5. x = phase delay, y = cost-to-fix rank. Plots use random samples from each phase for readability.}
\label{fig:phase_fixtime_plot}
\end{figure*}

@@
